Modelling low Mach number stellar hydrodynamics with MAESTROeX

### 2019-10-03 astronum_2019_MAESTROeX/astronum-maestroex.tex

The low Mach number fluid equations are given by 
\begin{align}
    \pd{\left(\rho X_k\right)}{t} &= - \nabla\cdot\left(\rho X_k \Ub \right) + \rho\dot{\omega}_k, \\
    \pd{\Ub}{t} &= - \Ub\cdot\nabla\Ub - \frac{\beta_0}{\rho}\nabla\left(\frac{\pi}{\beta_0}\right) - \frac{\rho - \rho_0}{\rho} g \bm{e}_r,\\
    \pd{\left(\rho h\right)}{t} &= -\nabla\cdot\left(\rho h \Ub\right) + \md{p_0}{t} + \rho H_{\mathrm{nuc}},
\end{align}
where $\rho$, $\Ub$ and $h$ are the mass density, fluid velocity and specific enthalpy, $X_k$ and $\dot{\omega}_k$ are the species mass fraction and production rate of species $k$, and $H_{\mathrm{nuc}}$ is the energy release per time per unit mass. From the base state pressure, we can define a one-dimensional base state density, $\rho_0 = \rho_0(r, t)$, which represents the lateral average and is in hydrostatic equilibrium with $p_0$:
\begin{equation}
    \nabla p_0 = -\rho_0 g \bm{e}_r,
\end{equation}
where $g = g(r,t)$ is the magnitude of the gravitational acceleration, and $\bm{e}_r$ is the radial unit vector. The background stratification is captured by introducing a buoyancy-like term, $\beta_0$. It is defined as 
\begin{equation}
    \beta_0(r,t) = \rho_0(0,t) \exp\left(\int_0^r dr'\, \frac{1}{\overline{\Gamma}_1p_0}\pd{p_0}{r'} \right),
\end{equation}
where $\overline{\Gamma}_1$ is the lateral average of the first adiabatic exponent, $\Gamma_1 \equiv d(\ln p)/d(\ln \rho)|_s$, and $s$ is the entropy. The equations are closed by casting the equation of state (EoS) as a velocity divergence constraint. This is done by taking the Lagrangian derivative of the EoS for pressure as a function of the thermodynamic variables, substituting in the equations of motion for mass and energy, and requiring that the pressure is described by a function of $r$ and $t$ based on the condition of hydrostatic equilibrium. Details of this derivation can be found in \cite{Almgren2006a,Almgren2006b}. This constraint is given by 
\begin{equation}
    \nabla\cdot\left(\beta_0\Ub\right) = \beta_0 \left(S - \frac{1}{\overline{\Gamma}_1 p_0}\pd{p_0}{t} \right).
\end{equation}
Here, $S$ is an expansion term which describes local compressibility effects due to changes in composition and heating from reactions:
\begin{equation}
    S = -\sigma \sum_k \xi_k\dot{\omega}_k + \frac{1}{\rho p_\rho}\sum_k p_{X_k}\dot{\omega}_k + \sigma H_{\mathrm{nuc}},
\end{equation}
where we define the following thermodynamic quantities as
\begin{align*}
    p_{X_k} \equiv \left.\pd{p}{X_k}\right|_{\rho,T,X_{j,j\neq k}},\qquad
     \xi_k&\equiv \left.\pd{h}{X_k}\right|_{p, T,X_{j,j\neq k}},\qquad
     p_\rho\equiv \left.\pd{p}{\rho}\right|_{T, X_k},\\
     \sigma \equiv \frac{p_T}{\rho c_p p_\rho}, \qquad
     p_T&\equiv \left.\pd{p}{T}\right|_{\rho, X_k} \quad\mathrm{and}\quad 
     c_p\equiv \left.\pd{h}{T}\right|_{p, X_k}.
\end{align*}

>>>>

The low Mach number fluid equations are given by 
\begin{align}
    \pd{\left(\rho X_k\right)}{t} &= - \nabla\cdot\left(\rho X_k \Ub \right) + \rho\dot{\omega}_k, \\
    \pd{\Ub}{t} &= - \Ub\cdot\nabla\Ub - \frac{\beta_0}{\rho}\nabla\left(\frac{\pi}{\beta_0}\right) - \frac{\rho - \rho_0}{\rho} g \bm{e}_r,\\
    \pd{\left(\rho h\right)}{t} &= -\nabla\cdot\left(\rho h \Ub\right) + \md{p_0}{t} + \rho H_{\mathrm{nuc}},
\end{align}
where $\rho$, $\Ub$ and $h$ are the mass density, fluid velocity and specific enthalpy, $X_k$ and $\dot{\omega}_k$ are the species mass fraction and production rate of species $k$, and $H_{\mathrm{nuc}}$ is the energy release per time per unit mass. \maestroex~defines the base state pressure $p_0$ to be consistent with the one-dimensional base state density, $\rho_0 = \rho_0(r, t)$, which represents the lateral average and is in hydrostatic equilibrium with $p_0$:
\begin{equation}
    \nabla p_0 = -\rho_0 g \bm{e}_r,
\end{equation}
where $g = g(r,t)$ is the magnitude of the gravitational acceleration, and $\bm{e}_r$ is the radial unit vector. The background stratification is captured by introducing a buoyancy-like term, $\beta_0$. It is defined as 
\begin{equation}
    \beta_0(r,t) = \rho_0(0,t) \exp\left(\int_0^r dr'\, \frac{1}{\overline{\Gamma}_1p_0}\pd{p_0}{r'} \right),
\end{equation}
where $\overline{\Gamma}_1$ is the lateral average of the first adiabatic exponent, $\Gamma_1 \equiv d(\ln p)/d(\ln \rho)|_s$, and $s$ is the entropy. The equations are closed by casting the equation of state (EoS) as a velocity divergence constraint. This is done by taking the Lagrangian derivative of the EoS for pressure as a function of the thermodynamic variables, substituting in the equations of motion for mass and energy, and requiring that the pressure is described by a function of $r$ and $t$ based on the condition of hydrostatic equilibrium. Details of this derivation can be found in \cite{Almgren2006a,Almgren2006b}. This constraint is given by 
\begin{equation}
    \nabla\cdot\left(\beta_0\Ub\right) = \beta_0 \left(S - \frac{1}{\overline{\Gamma}_1 p_0}\pd{p_0}{t} \right).
\end{equation}
Here, $S$ is an expansion term which describes local compressibility effects due to changes in composition and heating from reactions:
\begin{equation}
    S = -\sigma \sum_k \xi_k\dot{\omega}_k + \frac{1}{\rho p_\rho}\sum_k p_{X_k}\dot{\omega}_k + \sigma H_{\mathrm{nuc}},
\end{equation}
where we define the following thermodynamic quantities as
\begin{align*}
    p_{X_k} \equiv \left.\pd{p}{X_k}\right|_{\rho,T,X_{j,j\neq k}},\qquad
     \xi_k&\equiv \left.\pd{h}{X_k}\right|_{p, T,X_{j,j\neq k}},\qquad
     p_\rho\equiv \left.\pd{p}{\rho}\right|_{T, X_k},\\
     \sigma \equiv \frac{p_T}{\rho c_p p_\rho}, \qquad
     p_T&\equiv \left.\pd{p}{T}\right|_{\rho, X_k} \quad\mathrm{and}\quad 
     c_p\equiv \left.\pd{h}{T}\right|_{p, X_k}.
\end{align*}
